---
title: "Data on housing"
output:
  pdf_document: default
  html_notebook: default
---
```{r libraries, include=FALSE}
library(tidyverse)
library(RColorBrewer)
library("rgdal")
library(sf)
```

The aim was to get broad estimates of tenure at the county level and at the CBSA level.

Two main sources were used :  
- the [American Community Survey 5-Year Estimates](https://www.census.gov/programs-surveys/acs/) of the U.S Census Bureau  
- the [Picture of Subsidized Households](https://www.huduser.gov/portal/datasets/assthsg.html) survey of the US Department of Housing and Urban Development  

### The American Community Survey
 
The 2012-2016 version of the American Community Survey (ACS) was used. [^1]
The variables of interest are the following :   
- occupied housing units by tenure  
- total population in occupied housing units by tenure   

[^1]: The 2012-2016 American Community Survey (ACS) data generally reflect the February 2013 Office of Management and Budget (OMB) definitions of metropolitan and micropolitan statistical areas; in certain instances the names, codes, and boundaries of the principal cities shown in ACS tables may differ from the OMB definitions due to differences in the effective dates of the geographic entities. Estimates of urban and rural population, housing units, and characteristics reflect boundaries of urban areas defined based on Census 2010 data. As a result, data for urban and rural areas from the ACS do not necessarily reflect the results of ongoing urbanization.

### Picture of Subsidized Households

The survey was conducted in 2012. Even though data on multiple programs were available, only data on "Public Housing" were kept. 
 
The variable of interest are :  
- Number of units under contract for federal subsidy and available for occupancy  
- Occupied units as the % of units available  
- Total number of people  

## Data at the county level 

#### The American Community Survey

##### Ocupied housing units by tenure

This first data set provides information on tenure. 

```{r data_units_county, include=F}
data_unit_county <- read_csv2("/Users/Alice/University of Warwick/Rathelot, Roland - CB_IL/Alice/housing/data/ACS_16_5YR_B25003_county.csv", col_names=T)
colnames(data_unit_county)
```

```{r first changes to data_units_county}
data_unit_county <- data_unit_county %>%
  select(-c(starts_with("HD02_"))) %>%
  rename(
    total_units = HD01_VD01,
    total_units_owner = HD01_VD02,
    total_units_renter = HD01_VD03
    ) %>%
  mutate(
    pct_owner_unit = total_units_owner/total_units*100,   # percentage of owner occupied units
    pct_renter_unit = total_units_renter/total_units*100, # percentage of renter occupied units
    pct_tot_unit = pct_renter_unit + pct_owner_unit,      # check
    GEOID = str_sub(GEO.id,-5,-1)                         # for the join
  )
```

```{r display data_units_county}
head(data_unit_county)
```

##### Total population in ocupied housing by tenure 

This second data set provides information on the total population in occupied housing units by tenure by year householder moved into unit. However, only the broad estimates, which ignored the time dimension, were kept so as to remain consistent with the 2012-2016 data above. 


```{r data_pop_county, include=FALSE}
data_pop_county <- read_csv2("/Users/Alice/University of Warwick/Rathelot, Roland - CB_IL/Alice/housing/data/ACS_16_5YR_B25026_county.csv", col_names=T)
colnames(data_pop_county)
```

```{r first changes to data_pop_county}
data_pop_county <- data_pop_county %>%
  select(c(1:7),c(20), -c(starts_with("HD02_"))) %>%
  rename(
    total_pop = HD01_VD01,
    total_pop_owner = HD01_VD02,
    total_pop_renter = HD01_VD09
  ) %>%
  mutate(
    pct_owner_pop = total_pop_owner/total_pop*100,  # percentage of owner population
    pct_renter_pop= total_pop_renter/total_pop*100, # percentage of renter population
    pct_tot_pop =pct_renter_pop + pct_owner_pop,    # check
    GEOID = str_sub(GEO.id,-5,-1)                   # for the join
  )
```

```{r display data_pop_county}
head(data_pop_county)
```


```{r first join county}
data_ACS_county <- left_join(data_pop_county, data_unit_county, by = c("GEOID", "GEO.id2", "GEO.id", "GEO.display-label"))

head(data_ACS_county)
```

There are 3,144 counties in the US (including Alaska and Puerto Rico). In the data, there are 3,142 observations.   

#### Picture of Subsidized Households

```{r PSH_county data load, include=FALSE}
data_county <- read_csv2("/Users/Alice/University of Warwick/Rathelot, Roland - CB_IL/Alice/housing/data/COUNTY_2012.csv", col_names=T)
```

```{r PSH county data transformation}
data_county <- data_county %>%
  filter(program_label=="Public Housing" & !(is.na(total_units))) %>%
  select(states, entities, total_units, pct_occupied, people_total, state) %>%
  rename(
    total_units_ph = total_units, 
    pct_occupied_ph = pct_occupied, 
    total_pop_ph = people_total
    ) %>%
  mutate(GEOID = str_sub(entities,-5,-1)) %>%
  filter(!(GEOID=="78999") & !(GEOID=="66999")) # removes Guam and VI

head(data_county)
```
The data set is only made of 2,010 observations. 

#### Final data set for county level

```{r join county}
data_combined_county <- full_join(data_ACS_county, data_county, by = ("GEOID"))
```

```{r join county transformation, include=F}
data_combined_county <- data_combined_county %>% select(-c("pct_tot_pop","pct_tot_unit")) 
data_combined_county <- data_combined_county[,c(9,1:3,16,15,20,4:8,19,10:14,17:18)]
```  

```{r head join county}
head(data_combined_county)
```

## Data at the Core-base statistical areas level 

#### The American Community Survey
 
##### Ocupied housing units by tenure

```{r data_units_CBSA, include=F}
data_unit_CBSA <- read_csv2("/Users/Alice/University of Warwick/Rathelot, Roland - CB_IL/Alice/housing/data/ACS_16_5YR_B25003_CBSA.csv", col_names = T)

colnames(data_unit_CBSA)
```

```{r first changes to data_units}
data_unit_CBSA <- data_unit_CBSA %>%
  select(-c(starts_with("HD02_"))) %>%
  rename(
    total_units = HD01_VD01,
    total_units_owner = HD01_VD02,
    total_units_renter = HD01_VD03
  ) %>%
  mutate(
    pct_owner_unit = total_units_owner/total_units*100,   # percentage of owner occupied units
    pct_renter_unit = total_units_renter/total_units*100, # percentage of renter occupied units
    pct_tot_unit = pct_renter_unit + pct_owner_unit,      # check
    GEOID = str_sub(GEO.id,-5,-1)                         # for the join
  )

```

```{r display data_units}
head(data_unit_CBSA)
```

##### Total population in ocupied housing by tenure

```{r data_pop_CBSA, include=FALSE}
data_pop_CBSA <- read_csv2("/Users/Alice/University of Warwick/Rathelot, Roland - CB_IL/Alice/housing/data/ACS_16_5YR_B25033_CBSA.csv", col_names = T)
colnames(data_pop_CBSA)
```

```{r first changes to data_pop_CBSA}
data_pop_CBSA <- data_pop_CBSA %>%
  select(c(1:7),c(18), -c(starts_with("HD02_"))) %>%
  rename(
    total_pop = HD01_VD01,
    total_pop_owner = HD01_VD02,
    total_pop_renter = HD01_VD08
  ) %>%
  mutate(
    pct_owner_pop = total_pop_owner/total_pop*100,  # percentage of owner population
    pct_renter_pop= total_pop_renter/total_pop*100, # percentage of renter population
    pct_tot_pop =pct_renter_pop + pct_owner_pop,    # check
    GEOID = str_sub(GEO.id,-5,-1)                   # for the join
  )
```

```{r display data_pop_CBSA}
head(data_pop_CBSA)
```


```{r first join CBSA}
data_ACS_CBSA <- left_join(data_pop_CBSA, data_unit_CBSA, by = c("GEOID", "GEO.id2", "GEO.id", "GEO.display-label"))

head(data_ACS_CBSA)

```

#### Picture of Subsidized Households

```{r PSH_CBSA, include=F}
data_CBSA <- read_csv2("/Users/Alice/University of Warwick/Rathelot, Roland - CB_IL/Alice/housing/data/CBSA_2012.csv", col_names = T)
```

```{r PSH_CBSA first transformations }
data_CBSA <- data_CBSA %>%
  filter(program_label == "Public Housing" & !(is.na(total_units))) %>%
  select(entities, total_units, pct_occupied, people_total) %>%
  rename(
    total_units_ph = total_units, 
    pct_occupied_ph=pct_occupied, 
    total_pop_ph = people_total
  ) %>%
  mutate(GEOID = str_sub(entities,1,5)) %>%
  filter(!(GEOID=="78999") & !(GEOID=="66999")) # removes Guam and VI

head(data_CBSA)
```
There are 929 CBSA in the US (including Puerto Rico). Our data set is made of 798 CBSA. 

#### Final data set for county level

```{r join CBSA}
data_combined_CBSA <- full_join(data_ACS_CBSA, data_CBSA, by = ("GEOID"))
```

```{r join CBSA transformation, include=F}
data_combined_CBSA <- data_combined_CBSA %>% select(-c("pct_tot_pop","pct_tot_unit"))
data_combined_CBSA <- data_combined_CBSA[,c(9,1:3,15,4:8,18,10:14,16:17)]
```

```{r head join CBSA}
head(data_combined_CBSA)
```

