---
title: "Data on housing"
output:
  pdf_document: default
  html_notebook: default
---
```{r libraries, include=FALSE}
library(tidyverse)
library(RColorBrewer)
library("rgdal")
library(sf)
```

The aim was to get broad estimates of tenure at the county level and at the CBSA level.

Two main sources were used :  
- the [American Community Survey 5-Year Estimates](https://www.census.gov/programs-surveys/acs/) of the U.S Census Bureau  
- the [Picture of Subsidized Households](https://www.huduser.gov/portal/datasets/assthsg.html) survey of the US Department of Housing and Urban Development  

### The American Community Survey
 
The 2012-2016 version of the American Community Survey (ACS) was used. [^1]
The variables of interest are the following :   
- occupied housing units by tenure  
- total population in occupied housing units by tenure   

[^1]: The 2012-2016 American Community Survey (ACS) data generally reflect the February 2013 Office of Management and Budget (OMB) definitions of metropolitan and micropolitan statistical areas; in certain instances the names, codes, and boundaries of the principal cities shown in ACS tables may differ from the OMB definitions due to differences in the effective dates of the geographic entities. Estimates of urban and rural population, housing units, and characteristics reflect boundaries of urban areas defined based on Census 2010 data. As a result, data for urban and rural areas from the ACS do not necessarily reflect the results of ongoing urbanization.

### Picture of Subsidized Households

The survey was conducted in 2012. Even though data on multiple programs were available, only data on "Public Housing" were kept. 
 
The variable of interest are :  
- Number of units under contract for federal subsidy and available for occupancy  
- Occupied units as the % of units available  
- Total number of people  

## Data at the county level 

#### The American Community Survey

##### Ocupied housing units by tenure

This first data set provides information on tenure. The available variables are :   
GEO.id2	Id2    
GEO.display-label	Geography  
HD01_VD01	Estimate; Total:  
HD02_VD01	Margin of Error; Total:  
HD01_VD02	Estimate; Total: - Owner occupied  
HD02_VD02	Margin of Error; Total: - Owner occupied  
HD01_VD03	Estimate; Total: - Renter occupied  
HD02_VD03	Margin of Error; Total: - Renter occupied 

I only used the estimates and ignored the margins of error. 
```{r data_units_county, include=F}
data_unit_county <- read_csv2("/Users/Alice/Documents/ENSAE/2A/Warwick/Housing data/ACS/ACS_16_5YR_B25003_with_ann.csv", col_names = T)
colnames(data_unit_county)
```

```{r first changes to data_units_county}
data_unit_county <- data_unit_county %>%
  mutate(
    pct_owner_unit = HD01_VD02 / HD01_VD01 * 100, # percentage of owner occupied units
    pct_renter_unit = HD01_VD03 / HD01_VD01 * 100, # percentage of renter occupied units
    total_unit = pct_renter_unit + pct_owner_unit, # check
    GEOID = str_sub(GEO.id, -5, -1), # for the join
  )
```

```{r display data_units_county}
head(data_unit_county)
```

##### Total population in ocupied housing by tenure 

This second data set provides information on the total population in occupied housing units by tenure by year householder moved into unit. However, only the broad estimates, which ignored the time dimension, were kept so as to remain consistent with the 2012-2016 data above. 

The variables of interest are :    
GEO.id2	Id2  
GEO.display-label	Geography  
HD01_VD01	Estimate; Total population in occupied housing units  
HD02_VD01	Margin of Error; Total population in occupied housing units  
HD01_VD02	Estimate; Owner occupied   
HD02_VD02	Margin of Error; Owner occupied  
HD01_VD09	Estimate; Renter occupied  
HD02_VD09	Margin of Error; Renter occupied  

Again, the "margins of error" were not taken into account. 

```{r data_pop_county, include=FALSE}
data_pop_county <- read_csv2("/Users/Alice/Documents/ENSAE/2A/Warwick/Housing data/ACS/ACS_16_5YR_B25026_with_ann.csv", col_names = T)
colnames(data_pop_county)
```

```{r first changes to data_pop_county}
data_pop_county <- data_pop_county %>%
  select(c(1:7), c(20:21)) %>%
  mutate(
    pct_owner_pop = HD01_VD02 / HD01_VD01 * 100, # percentage of owner population
    pct_renter_pop = HD01_VD09 / HD01_VD01 * 100, # percentage of renter population
    total_pop = pct_renter_pop + pct_owner_pop, # check
    GEOID = str_sub(GEO.id, -5, -1), # for the join
  )
```

```{r display data_pop_county}
head(data_pop_county)
```


```{r first join county}
data_ACS_county <- left_join(data_pop_county, data_unit_county, by = c("GEOID", "GEO.id2", "GEO.id", "GEO.display-label"))

head(data_ACS_county)
```

There are 3,144 counties in the US (including Alaska and Puerto Rico). In the data, there are 3,142 observations.   

#### Picture of Subsidized Households

```{r PSH_county data load, include=FALSE}
data_county <- read_csv2("/Users/Alice/Documents/ENSAE/2A/Warwick/Housing data/PSH/COUNTY_2012.csv", col_names = T)
```

```{r PSH_county data transformation}
data_county <- data_county %>%
  filter(program_label == "Public Housing" & !(is.na(total_units))) %>%
  select(states, entities, total_units, pct_occupied, people_per_unit, people_total, state) %>%
  mutate(
    GEOID = str_sub(entities, -5, -1)
  ) %>%
  filter(!(GEOID == "78999") & !(GEOID == "66999")) # removes Guam and VI

head(data_county)
```
The data set is only made of 2,010 observations. 

#### Final data set for county level

```{r join_county}
data_combined_county <- full_join(data_ACS_county, data_county, by = ("GEOID"))
head(data_combined_county)
```

## Data at the Core-base statistical areas level 

#### The American Community Survey
 
##### Ocupied housing units by tenure

```{r data_units_CBSA, include=F}
data_unit_CBSA <- read_csv2("/Users/Alice/Documents/ENSAE/2A/Warwick/Housing data/ACS_CBSA/ACS_16_5YR_B25003_with_ann.csv", col_names = T)

colnames(data_unit_CBSA)
```

```{r first changes to data_units}
data_unit_CBSA <- data_unit_CBSA %>%
  mutate(
    pct_owner_unit = HD01_VD02 / HD01_VD01 * 100, # percentage of owner occupied units
    pct_renter_unit = HD01_VD03 / HD01_VD01 * 100, # percentage of renter occupied units
    total_unit = pct_renter_unit + pct_owner_unit, # check
    GEOID = str_sub(GEO.id, -5, -1), # for the join
  )
```

```{r display data_units}
head(data_unit_CBSA)
```

##### Total population in ocupied housing by tenure

```{r data_pop_CBSA, include=FALSE}
data_pop_CBSA <- read_csv2("/Users/Alice/Documents/ENSAE/2A/Warwick/Housing data/ACS_CBSA/ACS_16_5YR_B25033_with_ann.csv", col_names = T)
colnames(data_pop_CBSA)
```

```{r first changes to data_pop_CBSA}
data_pop_CBSA <- data_pop_CBSA %>%
  select(c(1:7), c(20:21)) %>%
  mutate(
    pct_owner_pop = HD01_VD02 / HD01_VD01 * 100, # percentage of owner population
    pct_renter_pop = HD01_VD09 / HD01_VD01 * 100, # percentage of renter population
    total_pop = pct_renter_pop + pct_owner_pop, # check
    GEOID = str_sub(GEO.id, -5, -1), # for the join
  )
```

```{r display data_pop_CBSA}
head(data_pop_CBSA)
```


```{r first join CBSA}
data_ACS_CBSA <- left_join(data_pop_CBSA, data_unit_CBSA, by = c("GEOID", "GEO.id2", "GEO.id", "GEO.display-label"))

head(data_ACS_CBSA)
```

#### Picture of Subsidized Households

```{r PSH_CBSA, include=F}
data_CBSA <- read_csv2("/Users/Alice/Documents/ENSAE/2A/Warwick/Housing data/PSH/CBSA_2012.csv", col_names = T)
```

```{r PSH_CBSA first transformations }
data_CBSA <- data_CBSA %>%
  filter(program_label == "Public Housing" & !(is.na(total_units))) %>%
  select(states, entities, total_units, pct_occupied, people_per_unit, people_total, state) %>%
  mutate(
    GEOID = str_sub(entities, -5, -1)
  ) %>%
  filter(!(GEOID == "78999") & !(GEOID == "66999")) # removes Guam and VI

head(data_CBSA)
```
There are 929 CBSA in the US (including Puerto Rico). Our data set is made of 798 CBSA. 

#### Final data set for county level

```{r join CBSA}
data_combined_CBSA <- full_join(data_ACS_CBSA, data_CBSA, by = ("GEOID"))
head(data_combined_CBSA)
```


